The \proglang{R} Package \pkg{forestinventory}: Design-Based Global and Small Area Estimations for Multiphase Forest Inventories

### 2017-08-11 JStatSoft_forestinventory_master.rnw
update Andreas 10.08.2017(1)
update Andreas 11.08.2017(2)

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Andreas Hill, Alexander Massey} %% comma-separated
\Plaintitle{A Capitalized Title: Something about a Package foo} %% without formatting
\Shorttitle{\pkg{forestinventory}: Design-Based Global and Small Area Estimations} %% a short title (if necessary)

@@

% set global options for R-output:
<<echo=FALSE>>=
options(prompt = "R> ", continue = " + ")
@

@@

@@

\bibliography{bib/literature}\newpage\cleardoublepage

@@

% Hallo. Here comes a first try:
% 
% <<>>=
% library(forestinventory)
% op <- onephase(formula = tvol~1 ,data = grisons,
%               phase_id =list(phase.col = "phase_id_2p",terrgrid.id = 2))
% summary(op)
% @
% 
% This was the result
% 
% 
%

>>>>

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Andreas Hill, Alexander Massey} %% comma-separated
\Plaintitle{A Capitalized Title: Something about a Package foo} %% without formatting
\Shorttitle{\pkg{forestinventory}: Design-Based Global and Small Area Estimations in \proglang{R}} %% a short title (if necessary)

@@

% set global options for R-output:
<<echo=FALSE>>=
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
@

@@

\SweaveInput{rnw/Analysis and Visualization.Rnw}

@@

\bibliography{bib/literature}\newpage

@@

### 2017-08-11 rnw/Twophase_Estimators and their Application.Rnw
update Andreas 11.08.2017(2)

An extension to the so-far published estimators by Mandallaz is the consideration of a \textit{boundary adjustment} according to forest/non-forest decision on the support level, which can be optionally applied in the \code{twophase} and \code{theephase} function of the package. In forest inventories, the sample is often restricted to those sample locations located within the forest area. In case a consistent forest definition can be applied to both the $s_2$ and $s_1$ sample (e.g. by a ploygon forest mask layer), it might be desired to restrict the calculation of the explanatory variables to the forest area within the given support (figure \ref{fig:exh_nexh_bw}). This method was suggested by \citet{mandallaz2013b} and led to an improvement of the regression model accuracy. To  insure  an  unbiased  calculation  of  the mean of $\pmb{Z}$, the  respective mean have then to be calculated as the \textit{weighted} mean according to equation \ref{eq:wmeanZ}. The weight $w(x)$ is thereby equal to the percentage of forested area within the support of sample location $x$.

@@

The \textbf{point estimate} for exhaustive ($ex$) and non-exhaustive ($nex$) explanatory variables is then calculated according to equation \ref{eq:pointest_2p_reg_ex} and  \ref{eq:pointest_2p_reg_nex} respectively. Note that this form results under the particular case where the regression coefficients are derived using the data from the current inventory (i.e. \textbf{internal} model). In this case, the mean of the residuals $\frac{1}{n_2}\sum_{x\in{s_2}}R(x)$ is zero by construction and does not have to be added as is necessary for \textbf{external} models \citep{mandallaz2016}.

@@

\begin{subequations}\label{eq:varexternal_2p_reg}
\begin{align}
  \hat{\var}_{ext}(\hat{Y}_{reg2p,ex}) & = \frac{1}{n_2}\frac{1}{n_{2}-1}\sum_{x \in {s_2}}\Big(Y(x)-\bar{Y}\Big)^2 \label{eq:varexternal_2p_reg_ex} \\
  \hat{\var}_{ext}(\hat{Y}_{reg2p,nex}) & =
 \frac{1}{n_1}\frac{1}{n_2-1}\sum_{x\in{s_2}}(Y(x)-\bar{Y}_2)^2+
(1-\frac{n_2}{n_1})\frac{1}{n_2}\frac{1}{n_2-1}\sum_{x\in{s_2}}(R(x)-\bar{R})^2 \label{eq:varexternal_2p_reg_nex}
\end{align}
\end{subequations}

@@

\begin{subequations}\label{eq:var_2p_small_psmall}
\begin{align}
  \hat{\var}(\hat{Y}_{G,small,2p}) & := \bar{\pmb{Z}}_G^t\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}\bar{\pmb{Z}}_G 
    + \frac{1}{n_{2,G}}\frac{1}{n_{2,G}-1}\sum_{x \in s_{2,G}} \Big(\hat{R}(x)-\hat{\bar{R}}_{2,G}\Big)^2  \label{eq:var_2p_reg_small} \\
  \hat{\var}(\hat{Y}_{G,psmall,2p}) & := \hat{\bar{\pmb{Z}}}_G^t\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}\hat{\bar{\pmb{Z}}}_G
  + \hat{\pmb{\beta}}_{s_2}^t\hat{\Sigma}_{\hat{\bar{\pmb{Z}}}_G}\hat{\pmb{\beta}}_{s_2} 
  + \frac{1}{n_{2,G}}\frac{1}{n_{2,G}-1}\sum_{x \in s_{2,G}} \Big(\hat{R}(x)-\hat{\bar{R}}_{2,G}\Big)^2  \label{eq:var_2p_reg_psmall}
\end{align}
\end{subequations}

\begin{subequations}\label{eq:varext_2p_small_psmall}
\begin{align}
  \hat{\var}_{ext}(\hat{Y}_{G,small,2p}) & := \frac{1}{n_{2,G}}\frac{1}{n_{2,G}-1}\sum_{x \in s_{2,G}} \Big(\hat{R}(x)-\hat{\bar{R}}_{2,G}\Big)^2  \label{eq:varext_2p_reg_small} \\
  \hat{\var}_{ext}(\hat{Y}_{G,psmall,2p}) & := \frac{1}{n_{2,G}}\frac{1}{n_{2,G}-1}\sum_{x \in s_{2,G}} \Big(Y(x)-\bar{Y}_{2,G}\Big)^2 \nonumber \\ 
 &+ \Big(1-\frac{n_{2,G}}{n_{1,G}}\Big)\frac{1}{n_{2,G}}\frac{1}{n_{2,G}-1}\sum_{x \in s_{2,G}} \Big(\hat{R}(x)-\hat{\bar{R}}_{2,G}\Big)^2 \label{eq:varext_2p_reg_psmall}
\end{align}
\end{subequations}

@@

In the \textit{small} and \textit{psmall} estimator, the crucial zero mean residual property for the small area $G$ is forced by the bias adjustment terms (equation \ref{eq:pointest_2p_small}, \ref{eq:pointest_2p_psmall}). However, the equations would considerably simplify if we could mathematically insure that the prediction model residuals over the entire area $F$ and the small area $G$ are by definition both zero at the same time, i.e. $\int_F R(x)dx=\int_G R(x)dx=0$. This idea is realized in the \textbf{\textit{extended}} \textbf{synthetic} and \textbf{\textit{extended}} \textbf{pseudo synthetic estimator} (\textit{extsynth} and \textit{extpsynth}) by extending the vector of auxiliary information $\pmb{Z}}(x)$ by a binary categorical indicator variable which has the value \code{1} if the observation lies inside the small area $G$ and is otherwise set to \code{0}. The \textit{extended} auxiliary vector thus reads as $\pmb{\mathbb{Z}}^t(x)=(\pmb{Z}^t(x),I_G(x))$ and corresponds to the classical ANOVA case. Note that this adjusts the intercept of the regression model for all data points of small area $G$ and thus insures that the mean of the residuals $\frac{1}{n_{2,G}}\sum_{x\in{s_{2,G}}}\mathbb{R}(x)$ in $G$ are always zero by construction.\par

@@

\begin{equation}\label{eq:ext_estvarmatrix}
  \hat{\pmb{\Sigma}}_{\hat{\pmb{\theta}}_{s_2}}:=\Big(\frac{1}{n_2}\sum_{x\in{s_2}}\pmb{\mathbb{Z}}(x)\pmb{\mathbb{Z}}^t(x) \Big)^{-1}
  \Big(\frac{1}{n_2^2}\sum_{x\in{s_2}}\hat{\mathbb{R}}^2(x)\pmb{\mathbb{Z}}(x)\pmb{Z}(x)^t\Big)
  \Big(\frac{1}{n_2}\sum_{x\in{s_2}}\pmb{\mathbb{Z}}(x)\pmb{\mathbb{Z}}^t(x) \Big)^{-1}
\end{equation}

@@

\begin{subequations}\label{eq:ext_varexternal_2p_extsynth}
\begin{align}
  \hat{\var}_{ext}(\hat{Y}_{G,extsynth,2p}) & = \frac{1}{n_{2,G}} \frac{1}{n_{2,G}-1}\sum_{x\in{s_{2,G}}}(\mathbb{R}(x)-\bar{\mathbb{R}})^2 \label{eq:ext_varexternal_2p_extsynth} \\
  \hat{\var}_{ext}(\hat{Y}_{G,extpsynth,2p}) & =
 \frac{1}{n_{1,G}}\frac{1}{n_{2,G}-1}\sum_{x\in{s_{2,G}}}(Y(x)-\bar{Y}_{2,G})^2 \nonumber \\ 
 &+ (1-\frac{n_{2,G}}{n_{1,G}})\frac{1}{n_{2,G}}\frac{1}{n_{2,G}-1}\sum_{x\in{s_{2,G}}}(\mathbb{R}(x)-\bar{\mathbb{R}})^2 \label{eq:ext_varexternal_2p_extpsynth}
\end{align}
\end{subequations}

@@

The \textbf{exhaustive versions of the small area estimators} (equations \ref{eq:pointest_2p_small},\ref{eq:var_2p_reg_small},\ref{eq:varext_2p_reg_small},\ref{eq:pointest_2p_reg_synth},\ref{eq:var_2p_reg_synth}) can again be applied by specifying the optional argument \code{exhaustive}. Its application requires that we know the exact means of all explanatory variables within the small area of interest. In contrast to the \textit{global} estimators, the exact means have now to be delivered in the form of a \code{data.frame}, where each row corresponds to a small area, and each column specifies the exact mean of the respective explanatory variable. Note that likewise the case of global estimation, the order of the explanatory variables in the data frame has to match the order in which they appear in the design matrix defined by the \code{lm} function in \proglang{R}. In order to tell \proglang{R} which row describes which small area, the row names have to match the respective names of the small areas specified in the \code{areas} argument.

>>>>

An extension to the so-far published estimators by Mandallaz is the consideration of a \textit{\textbf{boundary adjustment}} according to forest/non-forest decision on the support level, which can be optionally applied in the \code{twophase} and \code{theephase} function of the package. In forest inventories, the sample is often restricted to those sample locations located within the forest area. In case a consistent forest definition can be applied to both the $s_2$ and $s_1$ sample (e.g. by a ploygon forest mask layer), it might be desired to restrict the calculation of the explanatory variables to the forest area within the given support (figure \ref{fig:exh_nexh_bw}). This method was suggested by \citet{mandallaz2013b} and led to an improvement of the regression model accuracy. To  insure  an  unbiased  calculation  of  the mean of $\pmb{Z}$, the  respective mean have then to be calculated as the \textit{weighted} mean according to equation \ref{eq:wmeanZ}. The weight $w(x)$ is thereby equal to the percentage of forested area within the support of sample location $x$.

@@

The \textbf{point estimate} for exhaustive ($ex$) and non-exhaustive ($nex$) explanatory variables is then calculated according to equation \ref{eq:pointest_2p_reg_ex} and  \ref{eq:pointest_2p_reg_nex} respectively. Note that this form results under the particular case where the regression coefficients are derived using the data from the current inventory (i.e. \textbf{internal} model). In this case, the mean of the residuals $\frac{1}{n_2}\sum_{x\in{s_2}}\hat{R}(x)$ is zero by construction and does not have to be added as is necessary for \textbf{external} models \citep{mandallaz2016}.

@@

\begin{subequations}\label{eq:varexternal_2p_reg}
\begin{align}
  \hat{\var}_{ext}(\hat{Y}_{reg2p,ex}) & = \frac{1}{n_2}\frac{1}{n_{2}-1}\sum_{x \in {s_2}}\Big(Y(x)-\bar{Y}\Big)^2 \label{eq:varexternal_2p_reg_ex} \\
  \hat{\var}_{ext}(\hat{Y}_{reg2p,nex}) & =
 \frac{1}{n_1}\frac{1}{n_2-1}\sum_{x\in{s_2}}(Y(x)-\bar{Y}_2)^2+
(1-\frac{n_2}{n_1})\frac{1}{n_2}\frac{1}{n_2-1}\sum_{x\in{s_2}}(\hat{R}(x)-\bar{\hat{R}})^2 \label{eq:varexternal_2p_reg_nex}
\end{align}
\end{subequations}

@@

\begin{subequations}\label{eq:var_2p_small_psmall}
\begin{align}
  \hat{\var}(\hat{Y}_{G,small,2p}) & := \bar{\pmb{Z}}_G^t\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}\bar{\pmb{Z}}_G 
    + \frac{1}{n_{2,G}}\frac{1}{n_{2,G}-1}\sum_{x \in s_{2,G}} \Big(\hat{R}(x)-\bar{\hat{R}}_{2,G}\Big)^2  \label{eq:var_2p_reg_small} \\
  \hat{\var}(\hat{Y}_{G,psmall,2p}) & := \hat{\bar{\pmb{Z}}}_G^t\hat{\pmb{\Sigma}}_{\hat{\pmb{\beta}}_{s_2}}\hat{\bar{\pmb{Z}}}_G
  + \hat{\pmb{\beta}}_{s_2}^t\hat{\Sigma}_{\hat{\bar{\pmb{Z}}}_G}\hat{\pmb{\beta}}_{s_2} 
  + \frac{1}{n_{2,G}}\frac{1}{n_{2,G}-1}\sum_{x \in s_{2,G}} \Big(\hat{R}(x)-\bar{\hat{R}}_{2,G}\Big)^2  \label{eq:var_2p_reg_psmall}
\end{align}
\end{subequations}

\begin{subequations}\label{eq:varext_2p_small_psmall}
\begin{align}
  \hat{\var}_{ext}(\hat{Y}_{G,small,2p}) & := \frac{1}{n_{2,G}}\frac{1}{n_{2,G}-1}\sum_{x \in s_{2,G}} \Big(\hat{R}(x)-\bar{\hat{R}}_{2,G}\Big)^2  \label{eq:varext_2p_reg_small} \\
  \hat{\var}_{ext}(\hat{Y}_{G,psmall,2p}) & := \frac{1}{n_{2,G}}\frac{1}{n_{2,G}-1}\sum_{x \in s_{2,G}} \Big(Y(x)-\bar{Y}_{2,G}\Big)^2 \nonumber \\ 
 &+ \Big(1-\frac{n_{2,G}}{n_{1,G}}\Big)\frac{1}{n_{2,G}}\frac{1}{n_{2,G}-1}\sum_{x \in s_{2,G}} \Big(\hat{R}(x)-\bar{\hat{R}}_{2,G}\Big)^2 \label{eq:varext_2p_reg_psmall}
\end{align}
\end{subequations}

@@

In the \textit{small} and \textit{psmall} estimator, the crucial zero mean residual property for the small area $G$ is forced by the bias adjustment terms (equation \ref{eq:pointest_2p_small}, \ref{eq:pointest_2p_psmall}). However, the equations would considerably simplify if we could mathematically insure that the prediction model residuals over the entire area $F$ and the small area $G$ are by definition both zero at the same time, i.e. $\int_F R(x)dx=\int_G R(x)dx=0$. This idea is realized in the \textbf{\textit{extended}} \textbf{synthetic} and \textbf{\textit{extended}} \textbf{pseudo synthetic estimator} (\textit{extsynth} and \textit{extpsynth}) by extending the vector of auxiliary information $\pmb{Z}}(x)$ by a binary categorical indicator variable $I_G(x)$ which takes the value \code{1} if the observation lies inside the small area $G$ and is otherwise set to \code{0}. The \textit{extended} auxiliary vector thus reads as $\pmb{\mathbb{Z}}^t(x)=(\pmb{Z}^t(x),I_G(x))$ and leads to a classical ANOVA regression model. Note that the intercept of the regression model is thus adjusted for all data points belonging to small area $G$ which insures that the mean of the residuals $\frac{1}{n_{2,G}}\sum_{x\in{s_{2,G}}}R(x)$ in $G$ are always zero by construction.\par

@@

\begin{equation}\label{eq:ext_estvarmatrix}
  \hat{\pmb{\Sigma}}_{\hat{\pmb{\theta}}_{s_2}}:=\Big(\frac{1}{n_2}\sum_{x\in{s_2}}\pmb{\mathbb{Z}}(x)\pmb{\mathbb{Z}}^t(x) \Big)^{-1}
  \Big(\frac{1}{n_2^2}\sum_{x\in{s_2}}\hat{R}^2(x)\pmb{\mathbb{Z}}(x)\pmb{Z}(x)^t\Big)
  \Big(\frac{1}{n_2}\sum_{x\in{s_2}}\pmb{\mathbb{Z}}(x)\pmb{\mathbb{Z}}^t(x) \Big)^{-1}
\end{equation}

@@

\begin{subequations}\label{eq:ext_varexternal_2p_extsynth}
\begin{align}
  \hat{\var}_{ext}(\hat{Y}_{G,extsynth,2p}) & = \frac{1}{n_{2,G}} \frac{1}{n_{2,G}-1}\sum_{x\in{s_{2,G}}}(\hat{R}(x)-\bar{\hat{R}})^2 \label{eq:ext_varexternal_2p_extsynth} \\
  \hat{\var}_{ext}(\hat{Y}_{G,extpsynth,2p}) & =
 \frac{1}{n_{1,G}}\frac{1}{n_{2,G}-1}\sum_{x\in{s_{2,G}}}(Y(x)-\bar{Y}_{2,G})^2 \nonumber \\ 
 &+ (1-\frac{n_{2,G}}{n_{1,G}})\frac{1}{n_{2,G}}\frac{1}{n_{2,G}-1}\sum_{x\in{s_{2,G}}}(\hat{R}(x)-\bar{\hat{R}})^2 \label{eq:ext_varexternal_2p_extpsynth}
\end{align}
\end{subequations}

@@

The \textbf{exhaustive versions} of the small area estimators (equations \ref{eq:pointest_2p_small},\ref{eq:var_2p_reg_small},\ref{eq:varext_2p_reg_small},\ref{eq:pointest_2p_reg_synth},\ref{eq:var_2p_reg_synth}) can again be applied by specifying the optional argument \code{exhaustive}. Its application requires that we know the exact means of all explanatory variables within the small area of interest. In contrast to the \textit{global} estimators, the exact means have now to be delivered in the form of a \code{data.frame}, where each row corresponds to a small area, and each column specifies the exact mean of the respective explanatory variable. Note that likewise the case of global estimation, the order of the explanatory variables in the data frame has to match the order in which they appear in the design matrix defined by the \code{lm} function in \proglang{R}. In order to tell \proglang{R} which row describes which small area, the row names have to match the respective names of the small areas specified in the \code{areas} argument.
